Harmony: A Verifiable Systems Programming Language

### 2022-01-14 paper/paper.tex
More focus on systems programming

\emph{Harmony} is a Python-like programming language designed to teach
systems programming and verification techniques to students who do not have a
background in formal methods.  Harmony programs, which can be multi-threaded
or distributed, are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, trying to multiply two lists
with one another, and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\item Violations of interrupt safety: a race condition caused by the execution
of an interrupt handler.
\end{compactitem}

@@

Harmony's unique combination of features are a familiar and easy
to learn high-level programming language, the ability to concisely
specify an interface, the ability to compare behaviors of specification
and implementation, the ability to modularly model check programs,
and support for continuations and interrupts.  In addition, Harmony
finds various common problems automatically without a complicated
property specification language, such as Harmony's ability to find
instances of undesirable busy waiting.

@@

The Harmony distribution comes with a collection of specifications,
each often with multiple implementations.  The lock specification
comes with the spinlock implementation of \autoref{fig:synch}(a), but
also a ticket lock implementation and one based on continuations and
a ready queue.  The distribution also includes various concurrent
queue implementations, including the lock-free implementation
of~\cite{MS96}.

@@

@@

Each transition modifies at least the context bag and set of active
contexts.
A few operations (\texttt{Store}, \texttt{Stop}, and \texttt{Del})
also modify one of the shared variables.

@@

\paragraph{Frame/Return.}  \texttt{Frame} is the first operation of
a method and \texttt{Return} its last.  \texttt{Frame} pops a value of
the stack and assigns it to the argument (or argument list) of the
method.  It also saves the local variables onto the stack and sets the
value of the <{result}> variable to <{None}>.  The operation of
\texttt{Return} depends on whether the stack is empty of not.  If empty,
\texttt{Return} terminates the thread by removing the context from the
context bag and replacing the active set with the domain of the context bag.
If the stack is not empty, then (1) the saved local variables are popped
and restored, (2) the return address is popped and assigned to the
program counter, and (3) the value of <{result}> is pushed onto the
stack.

@@

\paragraph{SetIntLevel.}  A context contains a boolean that determines
if interrupts are enabled or not.  \texttt{SetIntLevel} pops the
new interrupt level and pushes the old.  Supporting interrupts is
important.  First, writing multi-threaded programs that also need
to handle interrupts is difficult.  Second, distributed algorithms
often require timer interrupts to trigger actions such as retransmissions.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{diners.png}
\end{center}
\caption{The HTML output of running the Dining Philosophers program}
\label{fig:dinersHTML}
\end{figure*}

>>>>

\emph{Harmony} is a Python-like programming language designed to teach
systems specification, implementation, and verification techniques to
students who do not have a background in formal methods.
Harmony programs, which can be multi-threaded
or distributed, are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, dereferencing dangling
pointers, trying to multiply two lists with one another,
and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\item Violations of interrupt safety: a race condition caused by the execution
of an interrupt handler.
\item Memory leaks.
\end{compactitem}

@@

Harmony's unique combination of features are a familiar and easy
to learn high-level programming language, the ability to concisely
specify an interface, the ability to compare behaviors of specification
and implementation, the ability to modularly model check programs,
and support for dynamic memory allocation, continuations, and
interrupts.  In addition, Harmony finds various common problems
automatically without a complicated property specification language,
such as Harmony's ability to find instances of undesirable busy
waiting.

@@

The Harmony distribution comes with a collection of specifications,
each often with multiple implementations.  The lock specification
comes with the spinlock implementation of \autoref{fig:synch}(a), but
also a ticket lock implementation and one based on continuations and
a blocked queue.  The distribution also includes various concurrent
queue implementations, including the lock-free implementation
of~\cite{MS96}.

@@

\begin{figure}
\begin{center}
{\footnotesize
\begin{code}
<{:inline:}>
import list

def Lock(held):
    result = { .held: False, .blocked: [] }

def acquire(lk):
    atomically:
        if lk->held:
            stop lk->blocked[len lk->blocked]
        else:
            lk->held = True

def release(lk):
    atomically:
        if lk->blocked == []:
            lk->held = False
        else:
            go (list.head(lk->blocked)) ()
            lk->blocked = list.tail(lk->blocked)
<{:end:}>
\end{code}
}
\end{center}
\caption{FIFO lock using continuation}
\label{fig:continuation}
\end{figure}

\subsection{Continuations}

The Harmony <{stop x}> expression stops the current thread and
saves its state in the variable <{x}>.  Another thread can
resume the thread using the statement <{go x r}>, which causes
the original thread to resume, with the <{stop x}> expression
evaluating to the value <{r}>.
\autoref{fig:continuation} shows another implementation of
\autoref{fig:synch}(a) using continuations, where blocked
threads are placed on a FIFO queue and released in that order.
Besides synchronization and scheduling, continuations also enable
the implementation of a primitive such as Posix \texttt{fork}.

\subsection{Interrupts}

Correct interrupt handling is another difficult aspect of
systems programming.  Interrupt handlers typically have various
stringent limitations in what they are allowed to do in order
to ensure the absence of race conditions due to re-entrant code.
The Harmony <{trap f(x)}> statement sets an interrupt handler
<{f(x}}> to be called at an unspecified later time.  You can think
of it as setting an unspecified timer, but it can model any type
of interrupt.  In addition, the Harmony <{setintlevel L}> statement
enables or disables interrupts and returns the old interrupt level.

@@

Each transition modifies at least the context bag and set of active
contexts.
A few operations (\texttt{Store}, \texttt{Stop}, and \texttt{Del})
also modify one of the shared variables.
Below we describe in more detail some of the more unusual ones.

@@

\begin{comment}
\paragraph{Frame/Return.}  \texttt{Frame} is the first operation of
a method and \texttt{Return} its last.  \texttt{Frame} pops a value of
the stack and assigns it to the argument (or argument list) of the
method.  It also saves the local variables onto the stack and sets the
value of the <{result}> variable to <{None}>.  The operation of
\texttt{Return} depends on whether the stack is empty of not.  If empty,
\texttt{Return} terminates the thread by removing the context from the
context bag and replacing the active set with the domain of the context bag.
If the stack is not empty, then (1) the saved local variables are popped
and restored, (2) the return address is popped and assigned to the
program counter, and (3) the value of <{result}> is pushed onto the
stack.
\end{comment}

@@

\begin{comment}
\paragraph{SetIntLevel.}  A context contains a boolean that determines
if interrupts are enabled or not.  \texttt{SetIntLevel} pops the
new interrupt level and pushes the old.  Supporting interrupts is
important.  First, writing multi-threaded programs that also need
to handle interrupts is difficult.  Second, distributed algorithms
often require timer interrupts to trigger actions such as retransmissions.
\end{comment}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=.8\textwidth]{diners.png}
\end{center}
\caption{The HTML output of running the Dining Philosophers program}
\label{fig:dinersHTML}
\end{figure*}
